---
title: "de Jong et al (2018)"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of N-pulse experiments

This document provides with the code to analyse data from two N-pulse experiments.

```{r, message = FALSE}
library(tidyverse)
library(patchwork)

# Load function used to annotate plots with correlation
source("./functions/corrLabel.R")

# Change ggplot2 default aesthetics
theme_set(theme_classic() + 
            theme(panel.grid = element_blank(), 
                  text = element_text(size = 8),
                  plot.title = element_text(hjust = -0.07, size = 10, face = "bold", 
                                            margin = margin(t = -5, b = 1)),
                  plot.subtitle = element_text(size = 10, hjust = 0.5)))
```


## Read data

Below, we read data from the pulse experiments, but also from the QTL experiments, 
so that they can be compared with each other. 

```{r, message = FALSE}
# Read pulse experiment data
pulse_plas <- read_csv("../../data/pulse/pulse_summary_wide.csv") %>% 
  rename_all(funs(str_replace(., "_mean_", "_pulse_")))

# Read MAGIC line data - for comparison with pulse experiment
magic_plas <- read_rds("../../data/phenotypes/magic_plasticity.rds") %>% 
  filter(id_kover %in% pulse_plas$id_public) %>% 
  select(id_kover, totalbr_mean_ln, totalbr_mean_hn, height_mean_ln, height_mean_hn, bolt_mean_ln, bolt_mean_hn) %>% 
  rename_all(funs(str_replace(., "_mean_", "_qtl_"))) %>% 
  rename(id_public = id_kover)

# Join to pulse data
## Shorten names of MAGIC lines for plotting
pulse_plas <- inner_join(pulse_plas, magic_plas, by = "id_public") %>% 
  mutate(id_public = str_replace(id_public, "MAGIC.", "M"))
```

## Fig S3 - Comparison between experiments

The following graph shows that the branching plasticity from this experiment is 
well correlated with that of the QTL experiment.

```{r}
# Get correlation for labeling plots
corr_labels <- pulse_plas %>% 
  group_by(experiment) %>% 
  summarise(x = -1.5,
            y = 7,
            corr_label = corLabel(totalbr_qtl_hn - totalbr_qtl_ln, totalbr_pulse_hn - totalbr_pulse_ln, TRUE))

p1 <- pulse_plas %>% 
  mutate(x = totalbr_qtl_hn - totalbr_qtl_ln,
         y = totalbr_pulse_hn - totalbr_pulse_ln) %>% 
  ggplot(aes(x, y)) +
  geom_abline(linetype = "dashed", colour = "grey") +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = id_public), size = 2.5) +
  geom_text(data = corr_labels, aes(label = corr_label), hjust = 0, size = 2.5) +
  scale_colour_manual(values = c("steelblue", "brown")) +
  labs(x = "Branching plasticity\n(QTL experiment)", y = "Branching plasticity\n(Pulse experiment)") +
  coord_cartesian(xlim = c(-1.5, 8), ylim = c(-1.5, 8))

#pdf("./figures/figureS3.pdf", width = 2.63, height = 2.63)
p1
#dev.off()
```

We can see that the flowering time was a bit different, where everyone flowered 
later than in the QTL experiment:

```{r}
pulse_plas %>% 
  ggplot(aes(bolt_pulse_ln - bolt_qtl_ln, bolt_pulse_hn - bolt_qtl_hn)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey") +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = id_public), size = 2.5) +
  scale_colour_manual(values = c("steelblue", "brown")) +
  labs(x = "LN\n(Pulse experiment - QTL experiment)", y = "HN\n(Pulse experiment - QTL experiment)",
       subtitle = "Flowering time difference between experiments") +
  theme(legend.position = "none")
```


# Fig 5 - Correlation between plasticity and response to a pulse

In both experiments there is a tentative positive, albeit statistically unconvincing, 
trend in that more plastic lines respond more to the pulse. 

```{r}
# Get correlation for labeling plots
corr_labels <- pulse_plas %>% 
  group_by(experiment) %>% 
  summarise(x = -1,
            y = 8,
            corr_label = corLabel(totalbr_pulse_pulse - totalbr_pulse_ln, totalbr_pulse_hn - totalbr_pulse_ln, TRUE))

# Make graph
#pdf("./figures/figure5.pdf", width = 3, height = 3)
pulse_plas %>% 
  mutate(x = totalbr_pulse_pulse - totalbr_pulse_ln,
         y =totalbr_pulse_hn - totalbr_pulse_ln) %>% 
  ggplot(aes(x, y)) +
  geom_abline(linetype = "dashed", colour = "grey") +
  geom_point() +
  geom_text(data = corr_labels, aes(label = corr_label), hjust = 0, size = 2.5) +
  ggrepel::geom_text_repel(aes(label = id_public), size = 2.5) +
  labs(x = "Branching plasticity\nPulse HN - LN", y = "Branching plasticity\nConstant HN - LN") +
  coord_equal(xlim = c(-1, 3), ylim = c(-1, 8)) +
  scale_x_continuous(breaks = c(0, 1, 2)) +
  scale_y_continuous(breaks = seq(0, 8, 2))
#dev.off()
```



# Session Info

```{r}
sessionInfo()
```

